Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-10 paper/paper.Rmd
More work on existing packages

@@

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}. Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984}. Consequently, the field of landscape ecology relies on tools to preprocess, modify, model, analyze and visualize spatial data.

@@

Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}. Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.

<!-- MH: Is this enough about the advantages of open-source? I feel like this could be a bit more in contrast to proprietary software, but also don't want to repeat stuff -->

@@

Even though many other open-source tools for landscape ecology exist \cite{Baker1992,QGISDevelopmentTeam2016,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language. For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}. Therefore, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology. In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge.

<!-- MH: Maybe there are better examples for open-source but not R than currently cited, e.g. GRASS -->

@@

@@

@@

## Spatial regression
<!-- MH: Do we really want to include this? -->

@@

@@

>>>>

<!-- MH: Not super happy with the list i) - iv) -->

@@

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}. Related to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}. Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}. Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}. Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and customize existing software to their specific needs easily \cite{Steiniger2009}.

@@

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language. Thus, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology. In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge. For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->

<!-- MH: Add references for all packages -->

@@

<!--MH: Short paragraph about scale; *grainchanger* https://doi.org/10.1111/2041-210X.13177 -->

@@

## Spatial regression
<!-- MH: Do we really want to include this? Not really a landscape ecology topic but rather (geo) spatial stats I would argue -->

@@

#### Connectivity {#sec:connectivity}
Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}. Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}. While structural connectivity focuses only on landscape characteristics (e.g. movement corridors, barriers, etc.), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.

Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}. On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}. Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}). Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}. In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}. The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks. More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.

<!-- MH: Work-in-Progress -->

<!-- - resistance surfaces (DOI: https://doi.org/10.1016/S0169-2046(02)00242-6): *gdistance* package DOI: 10.18637/jss.v076.i13-->

<!-- - *samc*: an R package for connectivity modeling with spatial absorbing Markov chains; DOI: https://doi.org/10.1111/ecog.04891 -->

@@

<!-- MH: Work-in-Progress -->
<!-- ISOLATION BY RESISTANCE; DOI: 10.1111/j.0014-3820.2006.tb00500.x -->

#### Neutral landscape models {#sec:NLM}

### 2020-07-10 paper/paper.tex
More work on existing packages

Open-source software includes all software which is released under a
license that allows to freely use, modify and distribute the software
\cite{St.Laurent2008}. Open-source software development has many
advantages, such as fast innovation, transparency and reliability as
well as longevity mainly due to many diverse contributors
\cite{vonKrogh2006,St.Laurent2008}.

@@

Even though many other open-source tools for landscape ecology exist
\cite{Baker1992,QGISDevelopmentTeam2016,Porta2017}, in this review we
are going to focus on tools implemented in the \emph{R} programming
language. For more general overviews see
e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
Therefore, in the first part we are going to try to give a comprehensive
overview over existing \emph{R} packages for landscape ecology. In the
second part, we will identify topics for which \emph{R} packages are
currently missing to the best of our knowledge.

@@

@@

>>>>

Software to manage and analyze data becomes increasingly important in
modern scientific research \cite{Wilson2014} and many scientific studies
would not have been possible without open-source software
\cite{Prlic2012}. Open-source software includes all software which is
released under a license that allows to freely use, modify and
distribute the software \cite{St.Laurent2008}. Open-source software
development has many advantages, such as fast innovation, transparency
and reliability as well as longevity mainly due to many diverse
contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use
of open-source software facilitates (computational) reproducibility and
can allow a better understanding of the methodology
\cite{Prlic2012,Powers2019}. Lastly, open-source software allows other
scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012}
and customize existing software to their specific needs easily
\cite{Steiniger2009}.

@@

Even though many other open-source tools for landscape ecology exist
\cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in
this review we are going to focus on tools implemented in the \emph{R}
programming language. Thus, in the first part we are going to try to
give a comprehensive overview over existing \emph{R} packages for
landscape ecology. In the second part, we will identify topics for which
\emph{R} packages are currently missing to the best of our knowledge.
For more general overviews see
e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

\hypertarget{sec:connectivity}{%
\paragraph{Connectivity}\label{sec:connectivity}}

Connectivity is one of the core elements of landscape structure
\cite{Taylor1993} and thus one of the core concepts of landscape ecology
\cite{With2019}. Landscape connectivity describes how landscape
characteristics facilitate or hinder the movement of species
\cite{Tischendorf2000} or other aspects of mobility, such as dispersal,
gene or nutrient flow \cite{With2019}. While structural connectivity
focuses only on landscape characteristics (e.g.~movement corridors,
barriers, etc.), functional connectivity also includes behavior
characteristics of the species \cite{Tischendorf2000,With2019}.

Given its wide concept, many different measures of connectivity exists
\cite{Kindlmann2008}. On patch level structural connectivity can be
measured using nearest-neighbor distances or characterizations of the
patch neighborhood (e.g.~amount of suitable habitat)
\cite{Kindlmann2008,With2019}. Such measures are provided within the
\emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Another way to describe connectivity is based on graph theory with the
advantages that also functional connectivity can be included
\cite{Kindlmann2008}. In graph theory \cite{Laita2011} landscapes are
described by nodes (i.e.~habitat patches) connected by and functional
connections called links \cite{Laita2011}. The \emph{grainscape} package
\cite{Chubaty2020} provides a tool to model connectivity based on
spatially explicit networks. More general, the \emph{igraph} packages
\cite{Csardi2006} provides functionality related to graph theory.

@@

\hypertarget{sec:NLM}{%
\paragraph{Neutral landscape models}\label{sec:NLM}}
